Social Visual Analytics: DevOps for Data Scientists

### 2015-01-30 conclusion.tex
yes

We described the motivation introducing web technologies
for searching, collaborating, recommending, publishing
reusable notebooks as part of the visual analytics process.
We implemented these ideas in the RCloud prototype.
Rcloud assists software development and release using
an approach similar to DevOps in web services.
Experience with the prototype provided evidence that
data science teams, and the organizations in which they work,
benefit from having such capabilities.

Future work will address cross-language development,
source code analysis, richer recommendation techniques,
and improvements to the HTML5 front end.

>>>>

We presented a case for an environment that supports the entire
visual analytics work process, from data acquisition and exploratory
data analysis through development and production deployment. One might
think of this approach as ``DevOps for data science.'' The environment
supports collaboration and communication with capabilities for
creating and sharing experimental workbooks, version control,
searching, recommending, annotating, and deploying experiments
as web sites and as reusable services.

We implemented these ideas in the RCloud prototype.
Experience with the prototype provides evidence that data
science teams, and the organizations in which they work,
benefit from having such capabilities.

Next steps are to support cross-language development, to incorporate
richer source code analysis and recommendation techniques, and
to make improvements in the HTML5 human interface.

### 2015-01-30 discussion.tex
yes

Our back-end environment assumes a high degree of trust between users.
Access control for information security is delegated to the host
operating system and web server. In practice, most operating
systems and web servers rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. 
Information security in RCloud should be re-examined, but more
sophisticated approaches such as information flow analysis are
still an active research area, and many subtle problems remain
\cite{Moore:2011:SAF}. We are hopeful that having a richer
environment for collaboration and information sharing will encourage
practical applications for new approaches to information security
in visual analytics.

>>>>

Our back-end environment assumes a high degree of trust between users.
Access control for information security is delegated to the host
operating system and web server. In practice, most operating
systems and web servers rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. 
Information security in RCloud should be re-examined, but more
sophisticated approaches such as information flow analysis are
still an active research area, and many problems have
been identified \cite{Moore:2011:SAF}. Having a richer
environment for collaboration and information sharing will encourage
practical applications for new approaches to information security
in visual analytics.

### 2015-01-30 introduction.tex
yes

>>>>

Over fifty years ago, John Tukey published ``The Future of Data Analysis''
in which he defined a remarkably prescient agenda for the future of
data analysis and statistical computing. His vision has been realized
through environments like S and R, that have encouraged the adoption of
statistical graphics, and robust techniques for practical data modeling.
The software people use today has fulfilled his vision for the practice
of data analysis by experts.

The field has now moved beyond the boundaries of what Tukey originally
foresaw. Both computational and networking capabilities today far exceed
what could have been imagined fifty years ago. Cheap, highly parallel
virtualized systems allow analysts to solve problems at a scale that
was once difficult to conceive. But platforms today also enable collaboration
and support complex tasks and work processes in ways that were also
hard to imagine in the past. We turn our attention to this opportunity.

### 2015-01-30 system.tex
more notes
yes

The internal computing infrastructure of organizations has changed
radically in the last fifteen years. The shift from large
servers toward scalable, lower-cost, distributed systems (``the cloud'')
led to a software ecosystem of processes distributed over a
network (often virtually defined) communicating via some
high-level protocol, usually HTTP. HTTP is dominant because web
browsers and servers are ubiquitous and available in almost
all hardware devices, from tiny sensors, to handheld devices and
laptops, to rack-mounted servers.

@@

We designed RCloud around the front-facing API, which provides roughly
one entry point to correspond with each requirement.

The center panel of the screen is for editing code and visualizations through
the notebook interface. Supplementary ``asset'' files which are part of the
notebook but not part of the notebook flow, are edited on the right.
Controls at the top of the screen allow the user to quickly run, view, 
and share the notebook. The call mechanism is the URL itself; arguments are
read from the URL and the final cell is the result. On the left are the 
browsing area for seeing other users' notebooks and for searching, as well
as a help system.

@@

The basic unit of operation in RCloud is a \emph{notebook}. A notebook
holds a sequence of \emph{cells}, each of which represents a
snippet of code or hypertext in Markdown. This is not a novel idea;
executable documents structured in this way are present in a wide
variety of other software systems, including Mathematica, IPython and Sage.

One of the main differences between RCloud and other systems is the
notion that notebooks are ``always deployed''. This improves the
ability to share and modify experiments and compose the results
without binding to specific versions of a notebook.
On the other hand, there are a variety of situations where it might
be necessary to name a specific version. Rather than force notebook
designers to decide which versions need to be preserved,
we embrace \emph{transparent} versioning.
This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
and VisTrails's version tree \cite{Callahan:2006:VVM}, where every change in the state of the system is tracked.

Implementing an entire version-control subsystem from scratch would
be a large undertaking. Fortunately, we can build RCloud on top of Github's
\emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
for creating simplified git repositories (the main limitation being the
use of text-only files in a single directory). The GitHub web-service
API provides most of the semantics we need for the versioning
portion of the storage back end: access to previous versions,
comments, starring, and forking.

Using GitHub was mainly a tactical decision to reduce our
time-to-prototype. At the same time, it had the beneficial effect of
allowing our backend to communicate with separate processes over HTTP
and JSON. This had the serendipitous advantage of quickly enabling
full search over RCloud notebooks, using Apache SOLR.
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
for small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was successful. Future projects can gain many
of the same benefits by adopting a similar strategy.

%% %
%% We needed version control over a directory, gist provides that,
%% wrapped over an HTTP interface.
%% %
%% Mainly a tactical decision to save time.
%% %
%% Nevertheless, availability of notebook data as (barely) plain text had good side
%% effects, like ease to build search.
%% %
%% Expect same with suggest.

@@

In RCloud, reputation and interest are a relationship between
\emph{notebooks} and \emph{users}, rather than a relationship between
user pairs. We decided on this approach because we expect typical
RCloud deployments to have relatively few users, but each user to have
written relatively many notebooks. Under that assumption, assigning
interest to users would not provide sufficiently ``high-resolution'' data.

@@

>>>>

\stephen{Let's work from front to back: from human interface to the
services and frameworks that support it.}

\stephen{I don't think we need to even make this argument, it seems so obvious.}
The internal computing infrastructure of organizations has changed
radically in the last fifteen years. The shift from large
servers toward scalable, lower-cost, distributed systems (``the cloud'')
led to a software ecosystem of processes distributed over a
network (often virtually defined) communicating via some
high-level protocol, usually HTTP. HTTP is dominant because web
browsers and servers are ubiquitous and available in almost
all hardware devices, from tiny sensors, to handheld devices and
laptops, to rack-mounted servers.

@@

%\stephen{What??}
%We designed RCloud around the front-facing API, which provides roughly
%one entry point to correspond with each requirement.

Figure~\ref{fig:something} shows the RCloud developer interface.
The center panel is a notebook for code and visualizations.
Code is edited in this panel, and visualizations are rendered.
Supplementary ``asset'' files, which are part of the notebook,
though not part of its executable flow, are edited on the right.
Controls at the top of the screen allow the user to run, view, and
share the notebook. The call mechanism is the URL itself; arguments are
read from the URL and the final cell is the result. On the left is a
browsing area for searching and seeing other users' notebooks, and a
help system.

@@

The main unit of computation in RCloud is a \emph{notebook}.
A notebook holds a sequence of \emph{cells}, each of which contains a
snippet of code or hypertext in Markdown. This is not a novel idea;
executable documents structured this way are present in many
other systems, including Mathematica, IPython and Sage.

One of the main contributions of RCloud is the notion that notebooks
are ``always deployed''. This improves the ability to share and modify
experiments and compose the results without binding to specific versions
of a notebook.
On the other hand, there are situations where it might be important to
be able to name a specific version. We do not expect designers to 
decide which versions need to be preserved, but we embrace
\emph{transparent} versioning.
This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
and VisTrails's version tree \cite{Callahan:2006:VVM}, where every change in the state of the system is tracked.

To implement this, it is fortunate that we can build RCloud on top of
Github's \emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
for creating simplified git repositories, the main limitation being the
restriction to text-only files in a single directory. The GitHub web-service
API provides most of the semantics we need for the versioning
portion of the storage back end: access to previous versions,
comments, starring, and forking.

Using GitHub for storage and versioning also has the advantage of allowing
us to use other Github capabilities that can be invoked by HTTP and JSON.
Particularly, we can easily provide full search, using Apache SOLR
\stephen{whatever that is!}
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
for small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was successful. Future projects can gain many
of the same benefits by adopting a similar strategy.

@@

In RCloud, reputation and interest are a relationship between
\emph{notebooks} and \emph{users}, rather than a relationship between
user pairs. We chose this approach because we expect initial 
RCloud deployments to have relatively few users, but some users will
relatively many notebooks. Under that assumption, assigning
interest to users would not provide sufficiently ``high-resolution'' data.

@@

\carlos{compare against https://www.opencpu.org, compare against
  IPython notebooks. Two-way communication? Actually describe what it
  is. IPython manipulate notebooks? Understand and contrast}
